### 2022-11-24 dissertation/chapters/03-heuristics.tex
✨ (chapter4): Review Done
✨ (final-review): Done up to chapter 3

\begin{quotation}
      \noindent ``It is not the strongest of the species that survives. It is also not the most intelligent that survives. It is the one that is the most adaptable to change.''
\end{quotation}
\begin{flushright}
      Charles Darwin
\end{flushright}

\noindent
Many different techniques have been used to train \acp{FFNN}~\cite{ref:kingma:2014}. Finding the best technique to use to train a \acs{FFNN} has been shown to be problem dependent in many cases~\cite{ref:kheiri:2017}. Every technique has its characteristics, constraints, advantages and disadvantages. At the time of writing, the majority of work that is published around the training of \acp{FFNN}, involves the use of gradient-based techniques~\cite{ref:nel:2021}. Gradient-based techniques are not without flaws and can, for example, yield slow convergence or get trapped in local optima~\cite{ref:mingguang:2009}. Other techniques have also been used to successfully train \acp{FFNN}, including meta-heuristics such as \acf{PSO}~\cite{ref:rakitianskaia:2012, ref:vanwyk:2014}, \acf{DE}~\cite{ref:espinal:2011} and \acfp{GA}~\cite{ref:gupta:1999}.

@@

      \item \textbf{Section~\ref{sec:heuristics:gd}} presents the reader with seven low-level, gradient-based \index{heuristic}heuristics, including \acf{SGD}, \acf{Momentum}, \acf{NAG}, \acf{Adagrad}, \acf{RMSProp}, \acf{Adadelta} and \acf{Adam}.

      \item \textbf{Section~\ref{sec:heuristics:mh}} presents the reader with three different population-baed \acp{MH}, including \acf{PSO}, \acf{DE} and \acfp{GA}.

      \item \textbf{Section~\ref{sec:heuristics:summary}} provides the reader with a brief summary of the chapter.
\end{itemize}

@@

\noindent
Optimisation problems come in a wide variety, and can be defined in terms of the number of variables used (uni- vs. multivariate), the number of objective functions used (single- vs. multi-objective), the degree of linearity (linear vs. quadratic/polynomial), the number of optima (uni- vs. multi-modal), the nature of the environment (static vs. dynamic), the types of variables used (separable vs. inseparable, discrete vs. continuous) and the set of constraints that the solution must adhere to (constrained vs unconstrained).

@@

As stated in Chapter~\ref{chap:anns}, the training of an \acp{FFNN} is a particular type of optimisation problem, where the goal is to find the configuration of \textit{weights}, such that the \acs{FFNN} yields output that minimises some loss function. The mechanism by which the optimal weights for a \acs{FFNN} is sough out, is executed by an \index{optimisation algorithm}optimisation algorithm known as a \index{heuristic}heuristic. Discussions on the details of heuristics are presented in the following sections.

@@

The term \textit{heuristic} comes from the Latin word \textit{heuristicus} which means ``to find out or discover''.~\citeauthor{ref:romanycia:1985}~\cite{ref:romanycia:1985} provides a complete study on the history and origins of the term \index{heuristic}\textit{heuristic}. From their research, a proposal is made to define \index{heuristic}heuristics in the context of \acf{AI}, as any device, be it a program, rule, piece of knowledge, which is added to a problem-solving system, in expectation that, on average, the performance will improve.

In the context of this dissertation, a \index{heuristic}heuristic refers to an algorithmic search technique that serves as a guide to a search process where good solutions to a optimisation problem is being sought out. Different \index{heuristic}heuristics make use of different information during the search process~\cite{ref:kheiri:2017}. During the training of \acp{FFNN}, \index{heuristic}heuristics such as gradient-based \index{heuristic}heuristics make use of the derivatives obtained by evaluating the \acs{FFNN}. It can thus be said that gradient-based \index{heuristic}heuristics make use of information directly from the \textit{search space}. On the contrary, \index{heuristic}heuristics such as \acp{MH} make use of meta-information obtained as a result of evaluating the \acs{FFNN}~\cite{ref:blum:2003}. The meta-information that \acp{MH} make use of could include ranked-performance of a population of candidate solutions, referred to as \textit{entities}. \index{meta-heuristic}Meta-heuristics are useful when there is imperfect information about the search space~\cite{ref:bianchi:2009}, and are generally less problem-specific than other classes of \index{heuristic}heuristics~\cite{ref:blum:2003}. This disseration takes a particular interest in gradient-based \index{heuristic}heuristics and \acp{MH}. Each of these are presented in detail in the following sections.

@@

Chapter~\ref{chap:anns} introduced the reader to \index {supervised learning}supervised learning and presented a number of loss functions. In the context of supervised learning, loss functions produce a scalar value $\epsilon$, that represents the error between the output of the \acs{FFNN} and the desired output. When using \acs{GD} to train \acp{FFNN}, the \index{loss function}loss function is used to adjust the weights of the \acs{FFNN} in order to minimise the error~\cite{ref:engelbrecht:2007}.~\citeauthor{ref:engelbrecht:2007}~\cite{ref:engelbrecht:2007} states that training of \acp{FFNN} using \acs{GD}, is done by calculating the gradient of $\epsilon$ in \textit{weight-space}, and then moving the weight vector along the negative gradient. An illustration of \acs{GD} is given in Figure~\ref{fig:heuristics:gd:gd_illustration}.

\begin{figure}[htbp]
      \centering
      \includegraphics[width=0.85\textwidth]{images/gradient_descent.pdf}
      \caption{An illustration of \acf{GD} over various timesteps showing the minimisation of the error with regards to weight value.}
      \label{fig:heuristics:gd:gd_illustration}
\end{figure}

\noindent
In the context of training \textit{shallow} \acp{FFNN} using \index{supervised learning}supervised learning, the error signal is propagated backwards in the network from the output layer, through the hidden layer to the input layer, updating the weights at the output and hidden layers. The algorithm that propagates the error signal backwards is known as \acf{BP}. \Acs{BP} was popularised by~\citeauthor{ref:werbos:1994}~\cite{ref:werbos:1994}.~\citeauthor{ref:nel:2021}~\cite{ref:nel:2021} states that \acs{BP} provides a procedure for updating the network weights, layer by layer, by evaluating the derivatives of the error function $\mathcal{E}$, with respect to the weights at each layer.~\citeauthor{ref:engelbrecht:2007}~\cite{ref:engelbrecht:2007} describes the \acs{BP} process in two steps:

@@

\noindent
The update step for \acs{GD} by means of \acs{BP} can be formulated as is shown in Equations~\eqref{eq:heuristics:gd:update_step_part_1}-\eqref{eq:heuristics:gd:update_step_part_3} below.

@@

\noindent
In Equation~\eqref{eq:heuristics:gd:update_step_part_1} above, $w_{i}$ is the weight value at index $i$, $t$ is the timestep, $\Delta w_{i}(t)$ is the weight update vector at index $i$ and timestep $t$. $\Delta w_{i}(t)$ is specified by Equation~\eqref{eq:heuristics:gd:update_step_part_2} below.

@@

\noindent
In Equation~\eqref{eq:heuristics:gd:update_step_part_2} above, $\eta$ is the learning rate and $\frac{\partial \epsilon}{\partial w_{i}}$ is the \textit{gradient} of the error, as a result of a loss function $\mathcal{E}$, relative to the weight at index $i$. The learning rate controls the step-size that is taken in the direction of the negative gradient at each timestep. Assuming the use of \acs{SSE} as the loss function, the partial derivative of $\epsilon$, relative to the weight $w$, at index $i$, is given by Equation~\eqref{eq:heuristics:gd:update_step_part_3} below.

@@

\noindent
In Equation~\eqref{eq:heuristics:gd:update_step_part_3} above, $t_{p}$ refers to the target value for pattern $p$, $o_{p}$ refers to the output value for pattern $p$, $f$ refers to the \index{activation function}activation function, $net_{p}$ refers to the \index{net input signal}net input signal for pattern $p$, and $z_{i,p}$ refers to the input value at index $i$ for pattern $p$.

Finally, assuming the use of \index{sigmoid}sigmoid activation function, the partial derivative of the activation function, relative to the net input signal is given in Equation~\eqref{eq:heuristics:gd:update_step_part_4} below.

@@

\noindent
Again, assuming the use of \acs{SSE} as the loss function, the pseudo-code implementation for the generic \acs{GD} algorithm is taken from~\citeauthor{ref:engelbrecht:2007}~\cite{ref:engelbrecht:2007} and is presented in Algorithm~\ref{algo:heuristics:gd:gd} below.

@@

\noindent
A more detailed breakdown of the steps for \acs{BP}, as presented by ~\citeauthor{ref:engelbrecht:2007}~\cite{ref:engelbrecht:2007}, is given in Appendix~\ref{app:backpropagation}.

A simplified notation is proposed from which the gradient-based heuristics can be compared. Consider a simplification of the update-step for \acs{SGD} as presented in Equations~\eqref{eq:heuristics:gd:update_step_part_1} and~\eqref{eq:heuristics:gd:update_step_part_2} where $\frac{\partial \epsilon}{\partial w_{i}}$ is simply represented as the gradient term $g$, resulting in a simplified weight update step as is shown in Equation~\eqref{eq:heuristics:gd:sgd} below.

\begin{equation}
      \label{eq:heuristics:gd:sgd}
      \begin{split}
            w = w - \eta g
      \end{split}
\end{equation}

\noindent
In Equation~\eqref{eq:heuristics:gd:sgd} above, $w$ refers to the weight vector, $\eta$ refers to the learning rate as before and $g$ refers to the gradient of the error function relative to weight vector. Note that, vector notation and some subscripts related to layers, indices, timesteps and training patterns are omitted for convenience.

@@

\begin{figure}[htbp]
      \centering
      \includegraphics[width=0.9\textwidth]{images/sgd.pdf}
      \caption{An illustration of \acf{SGD} fluctuations during training as taken from~\cite{ref:sgd:2006}.}
      \label{fig:heuristics:gd:sgd}
\end{figure}

\noindent
With batch training, all the training patterns are presented at once and the \acs{FFNN} converges to the global minimum of the loss function w.r.t $w$ (assuming the loss function is convex). However, this is computationally expensive and impractical to implement in many situations. While \acs{SGD} is able to jump out of local minima into new, potentially better local minima~\cite{ref:ruder:2016}, \acs{SGD} complicates convergence to the minima as \acs{SGD} can potentially keep overshooting better minima. By slowly decreasing the learning rate $\eta$ during training, the same convergence behaviour is achieved as with batch training.

A compromise to overshooting minima that \acs{SGD} suffers from, is to make use of mini-batches of training patterns where a small number of training patterns are presented to the \acs{FFNN} at once. This is referred to as \textit{mini-batch} training. Input patterns from mini-batches are approximations of the total population of training patterns. According to~\citeauthor{ref:ruder:2016}~\cite{ref:ruder:2016} this has two main advantages:

@@

\noindent
In the context of this dissertation, the implementation of \acs{SGD} refers to the mini-batch training implementation of \acs{GD}. Although mini-batch training does provide a compromise between stochastic and batch training, there are still a number of challenges faced by mini-batch training. These include:

@@

\noindent
Alternative variants have been proposed that lead to better control over the convergence characteristics caused by \acs{GD}. The first of these \acs{GD} variants include \acs{Momentum} and is presented in the following section.

@@

\Acs{Momentum} ~\cite{ref:qian:1999} is a variant of the \acs{SGD} that helps accelerate \acs{SGD} in the relevant direction, dampening oscillations.~\citeauthor{ref:ruder:2016}~\cite{ref:ruder:2016} mentions that this is done by adding a fraction $\alpha$, of the weight update vector of the previous timestep to the current weight update vector. An illustration of \acs{SGD} with and without momentum is given in Figure~\ref{fig:heuristics:gd:sgd_with_and_without_momentum}.

@@

\noindent
The accumulation of momentum is presented in Equation~\eqref{eq:heuristics:gd:momentum_part_1} below, while the update step is then amended in Equation~\eqref{eq:heuristics:gd:momentum_part_2} below.

\begin{equation}
      \label{eq:heuristics:gd:momentum_part_1}
      \begin{split}
            v = \alpha v - \eta g
      \end{split}
\end{equation}

\begin{equation}
      \label{eq:heuristics:gd:momentum_part_2}
      \begin{split}
            w = w + v
      \end{split}
\end{equation}

\noindent
By redefining the \acs{SGD} update steps as shown above in Equations~\eqref{eq:heuristics:gd:momentum_part_1} and~\eqref{eq:heuristics:gd:momentum_part_2}, the \acs{Momentum} heuristic allows for the increase of momentum for dimensions whose gradients point in the same direction, while simultaneously reducing momentum for dimensions whose gradients change direction, leading to faster convergence and less oscillation. The momentum term $\alpha$ is usually set to 0.9~\cite{ref:engelbrecht:2007, ref:ruder:2016}.

@@

\noindent
The difference in the update step for \acs{Momentum} and the update step for \acs{NAG} as developed by~\citeauthor{ref:sutskever:2013-2}~\cite{ref:sutskever:2013-2} is described by~\citeauthor{ref:ruder:2016}~\cite{ref:ruder:2016} as follows. \acs{Momentum} first calculates the current gradient, represented by the small blue vector in Figure~\ref{fig:heuristics:gd:nag} and then takes a large step in the direction of the updated accumulated gradient, presented by the big blue vector. In contrast, \acs{NAG} first takes a big step in the direction of the previous accumulated gradient presented by the brown vector. At this point the gradient is measured and \acs{NAG} makes a correction represented by the red vector. The complete update step is represented by the green vector. By anticipating approximate future positions of the weights, the weight update step as defined by \acs{NAG} controls the optimisation process from going too fast and results in increased responsiveness~\cite{ref:bengio:2013}. Along with the same velocity update step for \acs{Momentum}, as presented in Equation~\eqref{eq:heuristics:gd:momentum_part_1}, the \acs{NAG} weight update step is given in Equation~\eqref{eq:heuristics:gd:nag_part_2} below.

\begin{equation}
      \label{eq:heuristics:gd:nag_part_2}
      \begin{split}
            w = w + \alpha v - \eta g
      \end{split}
\end{equation}

@@

\Acf{Adagrad} is an adaptation of \acs{SGD} which implements a learning rate for every element in the weight vector and is developed by Duchi et al.~\cite{ref:duchi:2011}.~\citeauthor{ref:ruder:2016}~\cite{ref:ruder:2016} mentions that \acs{Adagrad} adapts the learning rate to the elements of the weight vector, by  performing small updates (i.e. low learning rates) for weights associated with frequently occurring features and larger updates (i.e. high learning rates) for weights associated with infrequent features. For this reason, \acs{Adagrad} is well suited for dealing with situations where training data is sparse.

In the \acs{GD} variants presented thus far, the weight updates have been applied by making use of the same learning rate $\eta$, for all elements of the weight update vector. \Acs{Adagrad} uses a different learning rate for every weight element $w_{i}$, at every timestep $t$. The weight update step for \acs{Adagrad} is presented in Equation~\eqref{eq:heuristics:gd:adagrad_part_1} below.

@@

\noindent
In Equation~\eqref{eq:heuristics:gd:adagrad_part_1} above, $w_{i}$ is the $i$-th element of the weight vector, $G_{t,ii} \in \mathbb{R}^{d \times d}$ is a diagonal matrix where each diagonal element $i,i$, is the sum of the squared gradients w.r.t. $w_{i}$, up to timestep $t$ and $\epsilon$ is a smoothing term that avoids division by zero and is set to an insignificant value such as $1 \times 10^{-8}$. Since $G_{t}$ contains the sum of the squares of the gradients along its diagonal, the weight update step can be vectorised and updated using the matrix-vector product, represented by $\odot$, between $G_{t}$ and $g_{t,i}$. The simplified update step as implemented by \acs{Adagrad} is given in Equation~\eqref{eq:heuristics:gd:adagrad_part_2} below.

\begin{equation}
      \label{eq:heuristics:gd:adagrad_part_2}
      \begin{split}
            w_{t+1} = w_{t} - \frac{\eta}{\sqrt{G_{t} + \epsilon}} \odot g_{t,i}
      \end{split}
\end{equation}

\noindent
\Ac{Adagrad}'s main benefit is that it does not require manual tuning of the learning rate $\eta$. However, a problem with \acs{Adagrad} is in the accumulation of squared gradients in the denominator, represented by $G_{t}$. Since every term that is added is positive, $G_{t}$ keeps growing, leading to a situation where the learning rate shrinks to the point that the heuristic is no longer able to learn.

@@

Zeiler~\cite{ref:zeiler:2012} presents \acs{Adadelta}, an improvement of \acs{Adagrad} that eliminates the accumulation of squared gradients in the denominator.~\citeauthor{ref:ruder:2016}~\cite{ref:ruder:2016} mentions that instead of accumulating all past squared gradients as in the case of \acs{Adagrad}, \acs{Adadelta} restricts the window of accumulation to a window with a fixed size $W$. However, storing $W$ previous squared gradients is very inefficient. Instead, the accumulation of gradients over $W$ steps is defined recursively as a decaying average of all past squared gradients. The moving average of squared gradients at timestep $t$, denoted by $E[g^{2}]_{t}$  depends on a fraction $\alpha$ of the previous average of squared gradients and the current gradient, similar to the update step for \acs{Momentum}. Also similar to \acs{Momentum}, $\alpha$ is set to 0.9. The update step for $E[g^{2}]_{t}$ is presented in Equation~\eqref{eq:heuristics:gd:adadelta_part_1} below.

\begin{equation}
      \label{eq:heuristics:gd:adadelta_part_1}
      \begin{split}
            E[g^{2}]_{t} = \alpha E[g^{2}]_{t - 1} + (1 - \alpha)g_{t}^{2}
      \end{split}
\end{equation}

\noindent
To illustrate the difference between \acs{Adagrad} and \acs{Adadelta}, the \acs{SGD} weight update step for \acs{Adagrad}, as presented in~\eqref{eq:heuristics:gd:adagrad_part_2} is rewritten such that the diagonal matrix $G_{t}$, is replaced with the moving average of squared gradients as presented in Equation~\eqref{eq:heuristics:gd:adadelta_part_2} below.

\begin{equation}
      \label{eq:heuristics:gd:adadelta_part_2}
      \begin{split}
            w_{t+1} = w_{t} - \frac{\eta}{\sqrt{E[g^{2}]_{t} + \epsilon}} g_{t}
      \end{split}
\end{equation}

\noindent
Since the denominator is the \acf{RMS} error criterion of the gradient, the criteria short-hand notation is used as is depicted in Equation~\eqref{eq:heuristics:gd:adadelta_part_3} below.

\begin{equation}
      \label{eq:heuristics:gd:adadelta_part_3}
      \begin{split}
            w_{t+1} = w_{t} - \frac{\eta}{RMS[g]_{t}} g_{t}
      \end{split}
\end{equation}

>>>>

\begin{quotation}
      ``It is not the strongest of the species that survives. It is also not the most intelligent that survives. It is the one that is the most adaptable to change.''
\end{quotation}
\begin{flushright}
      - Charles Darwin
\end{flushright}

Many different techniques have been used to train \acp{FFNN}~\cite{ref:kingma:2014}. Finding the best technique to use to train a \acs{FFNN} has been shown to be problem dependent in many cases~\cite{ref:kheiri:2017}. Every technique has its characteristics, constraints, advantages and disadvantages. At the time of writing, the majority of work that is published around the training of \acp{FFNN}, involves the use of gradient-based techniques~\cite{ref:nel:2021}. Gradient-based techniques are not without flaws and can, for example, yield slow convergence or get trapped in local optima~\cite{ref:mingguang:2009}. Other techniques have also been used to successfully train \acp{FFNN}, including meta-heuristics such as \acf{PSO}~\cite{ref:rakitianskaia:2012, ref:vanwyk:2014}, \acf{DE}~\cite{ref:espinal:2011} and \acfp{GA}~\cite{ref:gupta:1999}.

@@

      \item \textbf{Section~\ref{sec:heuristics:gd}} presents seven low-level, gradient-based \index{heuristic}heuristics, including \acf{SGD}, \acf{Momentum}, \acf{NAG}, \acf{Adagrad}, \acf{RMSProp}, \acf{Adadelta} and \acf{Adam}.

      \item \textbf{Section~\ref{sec:heuristics:mh}} presents three different population-based \acp{MH}, including \acf{PSO}, \acf{DE} and \acfp{GA}.

      \item \textbf{Section~\ref{sec:heuristics:summary}} provides a brief summary of the chapter.
\end{itemize}

@@

Optimisation problems come in a wide variety, and can be defined in terms of the number of variables used (uni- vs. multivariate), the number of objective functions used (single- vs. multi-objective), the degree of linearity (linear vs. quadratic/polynomial), the number of optima (uni- vs. multi-modal), the nature of the environment (static vs. dynamic), the types of variables used (separable vs. inseparable, discrete vs. continuous) and the set of constraints that the solution must adhere to (constrained vs unconstrained).

@@

As stated in Chapter~\ref{chap:anns}, the training of an \acp{FFNN} is a particular type of optimisation problem, where the goal is to find the configuration of \textit{weights}, such that the \acs{FFNN} yields output that minimises some loss function. The mechanism by which the optimal weights for a \acs{FFNN} is sough out, is executed by an \index{optimisation algorithm}optimisation algorithm known as a \index{heuristic}heuristic.

@@

The term \textit{heuristic} comes from the Latin word \textit{heuristicus} which means ``to find out or discover''.~\citeauthor{ref:romanycia:1985}~\cite{ref:romanycia:1985} provide a complete study on the history and origins of the term \index{heuristic}\textit{heuristic}. From their research, a proposal is made to define \index{heuristic}heuristics in the context of \acf{AI}, as any device, be it a program, rule, piece of knowledge, which is added to a problem-solving system, in expectation that, on average, the performance will improve.

In the context of this dissertation, a \index{heuristic}heuristic refers to an algorithmic search technique that serves as a guide to a search process where good solutions to a optimisation problem is being sought out. Different \index{heuristic}heuristics make use of different information during the search process~\cite{ref:kheiri:2017}. During training of \acp{FFNN}, \index{heuristic}heuristics such as gradient-based \index{heuristic}heuristics make use of the derivatives obtained by evaluating the \acs{FFNN}. It can thus be said that gradient-based \index{heuristic}heuristics make use of information directly from the \textit{search space}. On the contrary, \index{heuristic}heuristics such as \acp{MH} make use of meta-information obtained as a result of evaluating the \acs{FFNN}~\cite{ref:blum:2003}. The meta-information that \acp{MH} make use of could include ranked-performance of a population of candidate solutions, referred to as \textit{entities}. \index{meta-heuristic}Meta-heuristics are useful when there is imperfect information about the search space~\cite{ref:bianchi:2009}, and are generally less problem-specific than other classes of \index{heuristic}heuristics~\cite{ref:blum:2003}. This disseration takes a particular interest in gradient-based \index{heuristic}heuristics and \acp{MH}.

@@

Chapter~\ref{chap:anns} introduced \index {supervised learning}supervised learning and presented a number of loss functions. In the context of supervised learning, loss functions produce a scalar value $\epsilon$, that represents the error between the output of the \acs{FFNN} and the desired output. When using \acs{GD} to train \acp{FFNN}, the \index{loss function}loss function is used to adjust the weights of the \acs{FFNN} in order to minimise the error~\cite{ref:engelbrecht:2007}.~\citeauthor{ref:engelbrecht:2007}~\cite{ref:engelbrecht:2007} states that training of \acp{FFNN} using \acs{GD}, is done by calculating the gradient of $\epsilon$ in \textit{weight-space}, and then moving the weight vector along the negative gradient. An illustration of \acs{GD} is given in Figure~\ref{fig:heuristics:gd:gd_illustration}.

\begin{figure}[htbp]
      \centering
      \includegraphics[width=0.75\textwidth]{images/gradient_descent.pdf}
      \caption{An illustration of \acf{GD} over various timesteps showing the minimisation of the error with regards to weight value.}
      \label{fig:heuristics:gd:gd_illustration}
\end{figure}

In the context of training \textit{shallow} \acp{FFNN} using \index{supervised learning}supervised learning, the error signal is propagated backwards in the network from the output layer, through the hidden layer to the input layer, updating the weights at the output and hidden layers. The algorithm that propagates the error signal backwards is known as \acf{BP}. \Acs{BP} was popularised by~\citeauthor{ref:werbos:1994}~\cite{ref:werbos:1994}.~\citeauthor{ref:nel:2021}~\cite{ref:nel:2021} states that \acs{BP} provides a procedure for updating the network weights, layer by layer, by evaluating the derivatives of the error function $\mathcal{E}$, with respect to the weights at each layer.~\citeauthor{ref:engelbrecht:2007}~\cite{ref:engelbrecht:2007} describes the \acs{BP} process in two steps:

@@

The update step for \acs{GD} by means of \acs{BP} can be formulated as is shown in Equations~\eqref{eq:heuristics:gd:update_step_part_1}-\eqref{eq:heuristics:gd:update_step_part_4}. The general weight update step is given as

@@

where $w_{i}$ is the weight value at index $i$, $t$ is the timestep, $\Delta w_{i}(t)$ is the weight update vector at index $i$ and timestep $t$. The delta weight term $\Delta w_{i}(t)$, is given as

@@

where $\eta$ is the learning rate and $\frac{\partial \epsilon}{\partial w_{i}}$ is the \textit{gradient} of the error, as a result of a loss function $\mathcal{E}$, relative to the weight at index $i$. The learning rate controls the step-size that is taken in the direction of the negative gradient at each timestep. Assuming the use of \acs{SSE} as the loss function, the partial derivative of $\epsilon$, relative to the weight $w$, at index $i$, is given as

@@

where $t_{p}$ refers to the target value for pattern $p$, $o_{p}$ refers to the output value for pattern $p$, $f$ refers to the \index{activation function}activation function, $net_{p}$ refers to the \index{net input signal}net input signal for pattern $p$, and $z_{i,p}$ refers to the input value at index $i$ for pattern $p$.

Finally, assuming the use of the \index{sigmoid}sigmoid activation function, the partial derivative of the activation function, relative to the net input signal is given as

@@

Again, assuming the use of \acs{SSE} as the loss function, the pseudo-code implementation for the generic \acs{GD} algorithm is taken from~\citeauthor{ref:engelbrecht:2007}~\cite{ref:engelbrecht:2007} and is presented in Algorithm~\ref{algo:heuristics:gd:gd} below.

@@

A simplified notation is proposed from which the gradient-based heuristics can be compared. Consider a simplification of the update-step for \acs{SGD} as presented in Equations~\eqref{eq:heuristics:gd:update_step_part_1} and~\eqref{eq:heuristics:gd:update_step_part_2} where $\frac{\partial \epsilon}{\partial w_{i}}$ is simply represented as the gradient term $g$, resulting in a simplified weight update step and is given as

\begin{equation}
      \label{eq:heuristics:gd:sgd}
      \begin{split}
            \boldsymbol{w} = \boldsymbol{w} - \eta \boldsymbol{g}
      \end{split}
\end{equation}

where $\boldsymbol{w}$ refers to the weight vector, $\eta$ refers to the learning rate as before and $\boldsymbol{g}$ refers to the gradient of the error function relative to weight vector. Note that some subscripts related to layers, indices, timesteps and training patterns are omitted for convenience.

@@

\begin{figure}[htbp]
      \centering
      \includegraphics[width=0.8\textwidth]{images/sgd.pdf}
      \caption{An illustration of \acf{SGD} fluctuations during training as taken from~\cite{ref:sgd:2006}.}
      \label{fig:heuristics:gd:sgd}
\end{figure}

With batch training, all the training patterns are presented at once and the \acs{FFNN} converges to the global minimum of the loss function w.r.t $\boldsymbol{w}$ (assuming the loss function is convex). However, this is computationally expensive and impractical to implement in many situations. While \acs{SGD} is able to jump out of local minima into new, potentially better local minima~\cite{ref:ruder:2016}, \acs{SGD} complicates convergence to the minima as \acs{SGD} can potentially keep overshooting better minima. By slowly decreasing the learning rate $\eta$ during training, the same convergence behaviour is achieved as with batch training.

A compromise is to make use of mini-batches of training patterns, where a small number of training patterns are presented to the \acs{FFNN} at once. This is referred to as \textit{mini-batch} training. Input patterns from mini-batches are approximations of the total population of training patterns. According to~\citeauthor{ref:ruder:2016}~\cite{ref:ruder:2016} this has two main advantages:

@@

In the context of this dissertation, the implementation of \acs{SGD} refers to the mini-batch training implementation of \acs{GD}. Although mini-batch training does provide a compromise between stochastic and batch training, there are still a number of challenges faced by mini-batch training. These include:

@@

Alternative variants have been proposed that lead to better control over the convergence characteristics caused by \acs{GD}. The first of these \acs{GD} variants include \acs{Momentum}.

@@

\Acs{Momentum} ~\cite{ref:qian:1999} is a variant of \acs{SGD} that helps accelerate \acs{SGD} in the relevant direction, dampening oscillations.~\citeauthor{ref:ruder:2016}~\cite{ref:ruder:2016} mentions that this is done by adding a fraction $\alpha$, of the weight update vector of the previous timestep to the current weight update vector. An illustration of \acs{SGD} with and without momentum is given in Figure~\ref{fig:heuristics:gd:sgd_with_and_without_momentum}.

@@

The accumulation of momentum is presented in Equation~\eqref{eq:heuristics:gd:momentum_part_1}, while the update step is then amended in Equation~\eqref{eq:heuristics:gd:momentum_part_2}.

\begin{equation}
      \label{eq:heuristics:gd:momentum_part_1}
      \begin{split}
            \boldsymbol{v} = \alpha \boldsymbol{v} - \eta \boldsymbol{g}
      \end{split}
\end{equation}

\begin{equation}
      \label{eq:heuristics:gd:momentum_part_2}
      \begin{split}
            \boldsymbol{w} = \boldsymbol{w} + \boldsymbol{v}
      \end{split}
\end{equation}

By redefining the \acs{SGD} update steps as shown above in Equations~\eqref{eq:heuristics:gd:momentum_part_1} and~\eqref{eq:heuristics:gd:momentum_part_2}, the \acs{Momentum} heuristic allows for the increase of momentum for dimensions whose gradients point in the same direction, while simultaneously reducing momentum for dimensions whose gradients change direction, leading to faster convergence and less oscillation. The momentum term $\alpha$ is usually set to 0.9~\cite{ref:engelbrecht:2007, ref:ruder:2016}.

@@

The difference in the update step for \acs{Momentum} and the update step for \acs{NAG} as developed by~\citeauthor{ref:sutskever:2013-2}~\cite{ref:sutskever:2013-2} is described by~\citeauthor{ref:ruder:2016}~\cite{ref:ruder:2016} as follows. \acs{Momentum} first calculates the current gradient, represented by the small \textcolor{blue}{blue} vector in Figure~\ref{fig:heuristics:gd:nag} and then takes a large step in the direction of the updated accumulated gradient, presented by the big \textcolor{blue}{blue} vector. In contrast, \acs{NAG} first takes a big step in the direction of the previous accumulated gradient presented by the \textcolor{brown}{brown} vector. At this point the gradient is measured and \acs{NAG} makes a correction represented by the \textcolor{red}{red} vector. The complete update step is represented by the green vector. By anticipating approximate future positions of the weights, the weight update step as defined by \acs{NAG} controls the optimisation process from going too fast and results in increased responsiveness~\cite{ref:bengio:2013}. Along with the same velocity update step for \acs{Momentum}, as presented in Equation~\eqref{eq:heuristics:gd:momentum_part_1}, the \acs{NAG} weight update step is given as

\begin{equation}
      \label{eq:heuristics:gd:nag_part_2}
      \begin{split}
            \boldsymbol{w} = \boldsymbol{w} + \alpha \boldsymbol{v} - \eta \boldsymbol{g}
      \end{split}
\end{equation}

@@

\Acf{Adagrad} is an adaptation of \acs{SGD} that implements a learning rate for every element in the weight vector and is developed by Duchi et al.~\cite{ref:duchi:2011}.~\citeauthor{ref:ruder:2016}~\cite{ref:ruder:2016} mentions that \acs{Adagrad} adapts the learning rate to the elements of the weight vector, by  performing small updates (i.e. low learning rates) for weights associated with frequently occurring features and larger updates (i.e. high learning rates) for weights associated with infrequent features. For this reason, \acs{Adagrad} is well suited for dealing with situations where training data is sparse.

In the \acs{GD} variants presented thus far, the weight updates have been applied by making use of the same learning rate $\eta$, for all elements of the weight update vector. \Acs{Adagrad} uses a different learning rate for every weight element $w_{i}$, at every timestep $t$. The weight update step for \acs{Adagrad} is given as

@@

where $w_{i}$ is the $i$-th element of the weight vector, $G_{t,ii} \in \mathbb{R}^{d \times d}$ is a diagonal matrix where each diagonal element $i,i$, is the sum of the squared gradients w.r.t. $w_{i}$, up to timestep $t$ and $\epsilon$ is a smoothing term that avoids division by zero and is set to an insignificant value such as $1 \times 10^{-8}$. Since $\boldsymbol{G_{t}}$ contains the sum of the squares of the gradients along its diagonal, the weight update step can be vectorised and updated using the matrix-vector product, represented by $\odot$, between $G_{t,i}$ and $g_{t,i}$. The simplified update step as implemented by \acs{Adagrad} is given as

\begin{equation}
      \label{eq:heuristics:gd:adagrad_part_2}
      \begin{split}
            w_{t+1} = w_{t} - \frac{\eta}{\sqrt{G_{t,i} + \epsilon}} \odot g_{t,i}
      \end{split}
\end{equation}

\Ac{Adagrad}'s main benefit is that it does not require manual tuning of the learning rate $\eta$. However, a problem with \acs{Adagrad} is in the accumulation of squared gradients in the denominator, represented by $\boldsymbol{G_{t}}$. Since every term that is added is positive, $\boldsymbol{G_{t}}$ keeps growing, leading to a situation where the learning rate shrinks to the point that the heuristic is no longer able to learn.

@@

Zeiler~\cite{ref:zeiler:2012} presents \acs{Adadelta}, an improvement of \acs{Adagrad} that eliminates the accumulation of squared gradients in the denominator.~\citeauthor{ref:ruder:2016}~\cite{ref:ruder:2016} mentions that instead of accumulating all past squared gradients as in the case of \acs{Adagrad}, \acs{Adadelta} restricts the window of accumulation to a window with a fixed size $W$. However, storing $W$ previous squared gradients is very inefficient. Instead, the accumulation of gradients over $W$ steps is defined recursively as a decaying average of all past squared gradients. The moving average of squared gradients at timestep $t$, denoted by $E[\boldsymbol{g}^{2}]_{t}$  depends on a fraction $\alpha$ of the previous average of squared gradients and the current gradient, similar to the update step for \acs{Momentum}. Also similar to \acs{Momentum}, $\alpha$ is set to 0.9. The update step for $E[\boldsymbol{g}^{2}]_{t}$ is given as

\begin{equation}
      \label{eq:heuristics:gd:adadelta_part_1}
      \begin{split}
            E[\boldsymbol{g}^{2}]_{t} = \alpha E[\boldsymbol{g}^{2}]_{t - 1} + (1 - \alpha)\boldsymbol{g}_{t}^{2}
      \end{split}
\end{equation}

To illustrate the difference between \acs{Adagrad} and \acs{Adadelta}, the \acs{SGD} weight update step for \acs{Adagrad}, as presented in~\eqref{eq:heuristics:gd:adagrad_part_2} is rewritten such that the diagonal matrix $\boldsymbol{G_{t}}$, is replaced with the moving average of squared gradients and is given as

\begin{equation}
      \label{eq:heuristics:gd:adadelta_part_2}
      \begin{split}
            \boldsymbol{w}_{t+1} = \boldsymbol{w}_{t} - \frac{\eta}{\sqrt{E[\boldsymbol{g}^{2}]_{t} + \epsilon}} \boldsymbol{g}_{t}
      \end{split}
\end{equation}

Since the denominator is the \acf{RMS} error criterion of the gradient, the criteria short-hand notation is used and is given as

\begin{equation}
      \label{eq:heuristics:gd:adadelta_part_3}
      \begin{split}
            \boldsymbol{w}_{t+1} = \boldsymbol{w}_{t} - \frac{\eta}{RMS[\boldsymbol{g}]_{t}} \boldsymbol{g}_{t}
      \end{split}
\end{equation}
